---
title: "CML SNV mixed models"
output:
  html_document:
    code_folding: hide
    df_print: paged
    toc: yes
    toc_float:
      collapsed: yes
    toc_depth: 3
    theme: lumen
  pdf_document:
    toc: yes
    toc_depth: '3'
  word_document:
    toc: yes
    toc_depth: '3'
params:
  code: TRUE
  boot: TRUE
---

# Aim

This analysis is to explore the best SNV burden mixed model to fit the CML colony data

## Setup

```{r, include=FALSE}
knitr::opts_chunk$set(echo = params$code)
knitr::opts_chunk$set(message = params$code)
knitr::opts_chunk$set(warning = params$code)
```

### Libraries

```{r libraries}
knitr::opts_chunk$set(collapse = TRUE)
  
library(tidyverse)
library(magrittr)

library(ggbeeswarm)
library(ggtext)
library(ggeffects)
library(ggnewscale)
library(ggh4x)
library(scales)
library(patchwork)

library(lme4)
library(optimx)
library(dfoptim)
library(lmeresampler)

library(sjPlot)
library(performance)
library(parameters)

import::from(.from = "dplyr", count, select, rename, slice)
```

### Functions

```{r Functions}
# Add totals to df
addTotal <-
  function(df) {
    df %>% mutate(across(where(is.logical), as.character)) %>%
      bind_rows(summarise(.,
                          across(where(is.numeric), sum,  na.rm = T),
                          across(where(is.character), ~ "Total")))
  }

# Lmer optimiser QC

# allfit summary 
all_fit.summary <- function(all_fit) {
  broom.mixed::glance(all_fit) %>%  select(optimizer, AIC, BIC, NLL_rel)  %>%  arrange(NLL_rel)

}


# allfit compare estimated ranked by fit metrics
all_fit.compare <- function(all_fit) {
  Reduce(left_join, list(
    broom.mixed::glance(all_fit), 
    Reduce(bind_rows, list(broom.mixed::tidy(all_fit) %>% filter(effect == "fixed"),
    lapply(all_fit, function(x){ x@optinfo$conv$opt}) %>% unlist() %>% enframe(name = "optimizer", value = "estimate") %>% mutate(term = "lme4.optimizer_code"),
    lapply(all_fit, function(x){ as.numeric(isSingular(x))}) %>% unlist() %>% enframe(name = "optimizer", value = "estimate") %>% mutate(term = "isSingular"),
    lapply(all_fit, function(x){ as.numeric(check_convergence(x))}) %>% unlist() %>% enframe(name = "optimizer", value = "estimate") %>% mutate(term = "Converged")
  ))))  %>%
  ggplot(aes(reorder(optimizer, -NLL_rel), estimate, colour = optimizer)) +
  geom_point() +
  facet_grid(0 ~ term, scale = "free_x") +
  coord_flip() + guides(colour = "none") +  
  theme(axis.text.x = element_text(angle = 30, hjust = 1, vjust = 1))

}

#  Model check
model.check <- function(model) {
  n <- rlang::expr_name(rlang::enexpr(model))
  res <- enframe(c(
    "Formula" = unlist(str_split(as.character(model@call), pattern = "\\="))[2],
    "Observations" = nrow(model@frame),
    "REML" = isREML(model),
    "Optimizer" = model@optinfo$optimizer,
    "Singular" = isSingular(model),
    "Converged" = check_convergence(model)
  )) %>%  tab_df(title = paste0("Model ", n),
                 col.header = c("Test", "Value"))
  return(res)
}

# get model formula
get_formula <- function(model) {
  str_sub(unlist(str_split(as.character(model@call), pattern = "\\="))[2])

}

```

## Data

```{r Data }
per_sample_statistics.n834.df <- 
  read_csv("../data/per_sample_statistics.n834.csv")

driver.palette <- c(`Wild-type` = "#0072B5FF", Other ="#E18727FF",  `*BCR::ABL1*`="#BC3C29FF")
driver.palette1 <- c(WT = "#0072B5FF", Other ="#E18727FF",  BCR_ABL="#BC3C29FF")
driver.palette2 <- c(Wt = "#0072B5FF", Mt="#BC3C29FF")
driver.palette2b <- c(Absence = "#717D6E", Presence="#BC3C29FF")
```

### Setup analysis variables

```{r Setup analysis variables}
# BCR::ABL1-mutated vs non mutated
per_sample_statistics.n834.df %<>%
  mutate(BCR_ABL1 = ifelse(grepl("BCR::ABL1", driver3, perl = T), "Mt", "Wt"))

#  set Wt as ref
per_sample_statistics.n834.df %<>%  mutate(BCR_ABL1 = factor(BCR_ABL1, levels = c("Wt", "Mt")))


# BCR::ABL1-mutated vs Other driver vs wildtype
per_sample_statistics.n834.df %<>% mutate(driver2 = ifelse(driver3 == "WT", "WT", ifelse(
  grepl("BCR::ABL1", driver3, perl = T),
  "BCR_ABL",
  "Other"
)))

per_sample_statistics.n834.df %<>% mutate(driver2 = factor(driver2, levels = c("WT", "Other", "BCR_ABL")))

# Set batch variable as character
per_sample_statistics.n834.df %<>% mutate(library.cluster = as.character(library.cluster))
```

In total `r nrow(per_sample_statistics.n834.df)` samples passed QC across `r length(unique(per_sample_statistics.n834.df$Patient))` patients. Age (age_at_sample_exact) was defined as the count of completed years from birth at sampling and sample mutant status (BCR_ABL1) was defined as *BCR::ABL1* positive (Mt; n=`r sum(per_sample_statistics.n834.df$BCR_ABL1 == "Mt")`) or negative (Wt; n=`r sum(per_sample_statistics.n834.df$BCR_ABL1 == "Wt")`).

```{r}
per_sample_statistics.n834.df %>%  count(Patient, BCR_ABL1) %>% pivot_wider(names_from = BCR_ABL1, values_from = n) %>%  addTotal() %>%  tab_df()
```

### Add labels

```{r Add labels}
attr(per_sample_statistics.n834.df$Length, "label") <- "Mean telomere length (bp)"
attr(per_sample_statistics.n834.df$nsub_adj, "label") <- "SNV burden (depth adjusted)"
attr(per_sample_statistics.n834.df$age_at_sample_exact, "label") <- "Age (years)"
attr(per_sample_statistics.n834.df$BCR_ABL1, "label") <- "BCR::ABL1 status"
attr(per_sample_statistics.n834.df$driver2, "label") <- "Driver status"
```

## Use of mixed models

Linear mixed models implemented in the R package "lme4" were used due to the repeated measures at the patient-level in the data set. Models were fitted with default "lme4" parameters, if a model did not converge, lme4::allFit() was used to refit the model to all available optimisers (provide by the "lme4", "optimx", and "dfoptim" R packages), the best optimiser was selected from non-singular and converged refits with the highest negative log-likelihood. Only non-singular and converged models were considered for model selection using the Bayesian information criterion (BIC).

# SNV model analysis

To model the effect of *BCR::ABL1* fusion status on SNV burden (nsub_adj), we will build the model (accounting for batch) in a stepwise fashion:

1.  Confirm a *BCR::ABL* fusion status effect as a fixed effect only

2.  Test the addition of explanatory variables to the random effect configuration

## 0. Remove PD57332 samples due to lack of wild-type

```{r}
per_sample_statistics.n799.df <- per_sample_statistics.n834.df %>%  filter(Patient != "PD57332")
plot3a.per_sample_statistics.n834.df <- per_sample_statistics.n834.df
rm(per_sample_statistics.n834.df)
```

As PD57332 didn't have any wild-type colonies our estimations of SNV mutation burden (nsub_adj) were expected to be biased, therefore we removed this patient from SNV burden analysis, leaving `r nrow(per_sample_statistics.n799.df)` samples passed QC across `r length(unique(per_sample_statistics.n799.df$Patient))` patients. Age (age_at_sample_exact) was defined as the count of completed years from birth at sampling and sample mutant status (BCR_ABL1) was defined as *BCR::ABL1* positive (Mt; n=`r sum(per_sample_statistics.n799.df$BCR_ABL1 == "Mt")`) or negative (Wt; n=`r sum(per_sample_statistics.n799.df$BCR_ABL1 == "Wt")`).

```{r}
per_sample_statistics.n799.df %>%  count(Patient, BCR_ABL1) %>% pivot_wider(names_from = BCR_ABL1, values_from = n) %>%  addTotal() %>%  tab_df()
```

## 1. Is there a driver (2-level, BCR::ABL1 or non-BCR::ABL1) effect on SNV nsub_adj?

Using the subset dataset (n=`r nrow(per_sample_statistics.n799.df)`), we compare the inclusion of *BCR::ABL1* fusion status as a fixed effect to a null model "nsub_adj \~ age_at_sample_exact + (1 \| Patient)".

#### Null model

```{r}
SNV.lmer_0 <-
  lmer(
    nsub_adj ~  age_at_sample_exact + (1 | Patient),
    data = per_sample_statistics.n799.df , REML = F
  )

model.check(SNV.lmer_0)
```

#### Model 1 - driver as fixed effect

```{r}
SNV.lmer2_1 <-
  lmer(
    nsub_adj ~  age_at_sample_exact + BCR_ABL1 +  (1 | Patient),
    data = per_sample_statistics.n799.df, REML = F
  )

model.check(SNV.lmer2_1)
```

### Compare null to driver as fixed effect

Adding *BCR::ABL* fusion status as a fixed effect improves the model fit to the data (`r deparse(substitute(SNV.lmer2_1))`; BIC=`r round(BIC(SNV.lmer2_1),2)`) over the null model (`r deparse(substitute(SNV.lmer_0))`; BIC=`r round(BIC(SNV.lmer_0),2)`), confirming a *BCR::ABL* fusion status effect.

```{r}
tab_model(
  SNV.lmer_0,
  SNV.lmer2_1,
  show.se = T,
  show.stat = T,
  show.ci = F,
  show.p = F,
  p.style = "scientific",
  dv.labels = c("Baseline (SNV.lmer_0)", "Driver as fixed effect (SNV.lmer2_1)")
)
```

#### ANOVA

```{r}
broom::tidy(anova(SNV.lmer_0, SNV.lmer2_1)) %>%
mutate(p.value = format.pval(p.value, digits = 3, eps = 1e-50)) %>% tab_df()
```

## 2. Random effect configuration

We now want to improve the model, so we test whether "Patient" has an effect on the slope as well as the intercept in 3 models;

-   Model 2 - Age at sampling
-   Model 3 - *BCR::ABL* fusion status

#### Model 2

```{r}
SNV.lmer2_2 <-
  lmer(
    nsub_adj ~  age_at_sample_exact + BCR_ABL1 + (1 + age_at_sample_exact | Patient),
    data = per_sample_statistics.n799.df,
    REML = F
  )

model.check(SNV.lmer2_2)
```

##### Check optimisers

```{r}
SNV.lmer2_2.af <- allFit(SNV.lmer2_2)

all_fit.summary(SNV.lmer2_2.af)
```

##### Plot optimiser outputs

```{r fig.width = 10, fig.height = 4, out.width="100%"}
all_fit.compare(SNV.lmer2_2.af)
```

##### Select best optimiser

```{r}
SNV.lmer2_2 <- SNV.lmer2_2.af$Nelder_Mead

model.check(SNV.lmer2_2)
```

#### Model 3

```{r}
SNV.lmer2_3 <-
  lmer(
    nsub_adj ~  age_at_sample_exact + BCR_ABL1 + (1 + BCR_ABL1 | Patient),
    data = per_sample_statistics.n799.df, REML = F
  )

model.check(SNV.lmer2_3)
```

### Model comparison

-   SNV.lmer2_2: nsub_adj \~ age_at_sample_exact + BCR_ABL1 + (1 + age_at_sample_exact \| Patient)

-   SNV.lmer2_3: nsub_adj \~ age_at_sample_exact + BCR_ABL1 + (1 + BCR_ABL1 \| Patient)

Both models are improvements over the baseline model's fit to the data (`r deparse(substitute(SNV.lmer2_1))`; BIC=`r round(BIC(SNV.lmer2_1),2)`), with `r deparse(substitute(SNV.lmer2_3))` (BIC=`r round(BIC(SNV.lmer2_3),2)`) identified as the better model over `r deparse(substitute(SNV.lmer2_2))` (BIC=`r round(BIC(SNV.lmer2_2),2)`).

```{r}
tab_model(
  SNV.lmer2_1,
  SNV.lmer2_2,
  SNV.lmer2_3,
  show.se = T,
  show.stat = T,
  show.ci = F,
  show.p = F,
  p.style = "scientific",
  dv.labels = c(
    "Baseline (SNV.lmer2_1)",
    "Random age effect and intercept: (SNV.lmer2_2)",
    "Random driver effect and intercept: (SNV.lmer2_3)"
  )
)
```

#### ANOVA

SNV.lmer2_1 vs. SNV.lmer2_2

```{r}
broom::tidy(anova(SNV.lmer2_1, SNV.lmer2_2)) %>%
mutate(p.value = format.pval(p.value, digits = 3, eps = 1e-50)) %>% tab_df()
```

SNV.lmer2_1 vs. SNV.lmer2_3

```{r}
broom::tidy(anova(SNV.lmer2_1, SNV.lmer2_3)) %>%
mutate(p.value = format.pval(p.value, digits = 3, eps = 1e-50)) %>% tab_df()
```

### Further selection

As we assume that SNV burden intercepts at the patient level will be similar, can we improve the model by removing the random effect intercept?

```{r}
SNV.lmer2_3_0r <-
  lmer(
    nsub_adj ~  age_at_sample_exact + BCR_ABL1 + (0 + BCR_ABL1 | Patient),
    data = per_sample_statistics.n799.df, REML = F
  )

model.check(SNV.lmer2_3_0r)
```

Both models are functionally equivalent with identical BIC values, therefore we retain the model with the random intercept (SNV.lmer2_3).

```{r}
tab_model(
  SNV.lmer2_3,
  SNV.lmer2_3_0r,
  show.se = T,
  show.stat = T,
  show.ci = F,
  show.p = F,
  p.style = "scientific",
  dv.labels = c(
    "Random effect of driver status and intercept: (SNV.lmer2_3)",
    "Random effect of driver status: (SNV.lmer2_3_0r)"
  )
)
```

#### ANOVA

```{r}
broom::tidy(anova(SNV.lmer2_3, SNV.lmer2_3_0r)) %>% 
  mutate(p.value = format.pval(p.value, digits = 3, eps = 1e-50)) %>% tab_df()
```

## 3. Final model

#### Refit with REML

```{r}
SNV.lmer2_3.REML <-
  update(
    SNV.lmer2_3,
    REML = T,
    control = lmerControl(
      optimizer = SNV.lmer2_3@optinfo$optimizer,
      optCtrl = SNV.lmer2_3@optinfo$control
    )
  )

model.check(SNV.lmer2_3.REML)
```

#### Check model

```{r , fig.height = 15, fig.width = 10, out.width="100%"}
check_model(SNV.lmer2_3.REML)
```

### Model summary

```{r}
tab_model(
  SNV.lmer2_3.REML,
  show.se = T,
  show.stat = T,
  show.ci = F,
  show.p = F,
  p.style = "scientific"
)
```

### Bootstrapped fixed effects CIs

```{r Bootstrap, eval = params$boot}
set.seed(1234)

SNV.lmer2_3_par_boot <-
  bootstrap(
    SNV.lmer2_3.REML,
    .f = fixef,
    type = "parametric",
    B = 3000
  )
SNV.lmer2_3_par_boot
```

#### Restricted to first 1000/`r SNV.lmer2_3_par_boot$B` bootstrapped models

```{r, eval = params$boot}
SNV.lmer2_3_par_boot2 <- SNV.lmer2_3_par_boot
SNV.lmer2_3_par_boot2$replicates <- SNV.lmer2_3_par_boot$replicates[1:1000, ]
lmeresampler:::confint.lmeresamp(SNV.lmer2_3_par_boot2, type = "perc") %>% tab_df()
```

```{r, eval = params$boot}
keep.lgl.vec <- sapply(SNV.lmer2_3_par_boot$message, is.null) & sapply(SNV.lmer2_3_par_boot$warning, is.null) & sapply(SNV.lmer2_3_par_boot$error, is.null)
```

#### Restricted to `r sum(keep.lgl.vec)` converged non-singular bootstrapped models

```{r, eval = params$boot}
SNV.lmer2_3_par_boot3 <- SNV.lmer2_3_par_boot
SNV.lmer2_3_par_boot3$replicates <- SNV.lmer2_3_par_boot$replicates[keep.lgl.vec, ][1:1000,]
lmeresampler:::confint.lmeresamp(SNV.lmer2_3_par_boot3, type = "perc") %>% tab_df()
```

#### Plot model

```{r , fig.height = 10, fig.width = 14, out.width="100%", eval = params$boot}

d <-
  ggpredict(
    model = SNV.lmer2_3.REML,
    terms = c("age_at_sample_exact [0:90 by=5]", "BCR_ABL1"),
    ci.lvl = 0.95,
    type = "fe"
  )

d %>% as_tibble() %>%
  mutate(!!sym(get_legend_title(d)) := group) %>%
  ggplot(aes(
    x = x,
    y = predicted,
    group = !!sym(get_legend_title(d)),
  )) +
  geom_line(aes(colour = !!sym(get_legend_title(d)))) +
  geom_ribbon(aes(
    ymin = conf.low,
    ymax = conf.high,
    fill  = !!sym(get_legend_title(d))
  ), alpha = 0.2) +
  geom_quasirandom(
    data = attr(d, "rawdata") %>% mutate(!!sym(get_legend_title(d)) := group),
    aes(x = x, y = response, colour = group),
    alpha = 0.5
  ) +
  scale_colour_manual(name = "BCR::ABL1 status", values = driver.palette2) +
  scale_fill_manual(name = "BCR::ABL1 status", values = driver.palette2) +
  xlab(get_x_title(d)) +
  ylab(get_y_title(d)) +
  new_scale_fill() +
  geom_abline(
    data = as_tibble(list(
      a = c(95.6),
      b = c(17.0),
      `Reference trend` = c("Williams et al. 17bp/year")
    )),
    aes(
      intercept = a,
      slope = b,
      fill  = `Reference trend`
    ),
    lty = 2,
    show.legend = T
  )  +
  ggtitle(unlist(str_split(
    as.character(SNV.lmer2_3.REML@call), pattern = "\\="
  ))[2]) +
  theme_bw(base_size = 16) 
```

### Figure 3a

Number of SNVs carried by sequenced colonies, coloured by driver status (*BCR::ABL1*, Other driver and wild-type) and their relationship with age, with mixed model trend lines. Red trend lines for colonies which are *BCR::ABL1*-positive and grey for *BCR::ABL1*-negative colonies.

```{r , fig.height = 10, fig.width = 14, out.width="100%", eval = params$boot}

Figure_3a_v2 <-
d %>% as_tibble() %>%
  mutate(group = case_when(group == "Wt" ~"Absence", group == "Mt" ~"Presence")) %>% 
  mutate(`*BCR::ABL1* status` := group) %>%
  ggplot(aes(
    x = x,
    y = predicted,
    group = `*BCR::ABL1* status`,
  )) +
  geom_line(aes(colour = `*BCR::ABL1* status`)) +
  geom_ribbon(aes(
    ymin = conf.low,
    ymax = conf.high,
    fill  = `*BCR::ABL1* status`
  ), alpha = 0.2) +
  scale_colour_manual(name = "*BCR::ABL1* status", values = driver.palette2b) +
  scale_fill_manual(name = "*BCR::ABL1* status", values = driver.palette2b) +
  new_scale_fill() +
  new_scale_colour() +
  geom_quasirandom(
    data = plot3a.per_sample_statistics.n834.df %>%
    mutate(response = nsub_adj, x = age_at_sample_exact, group = driver2) %>% mutate(`Driver status` = factor(
      driver2,
      levels = c("BCR_ABL", "WT", "Other"),
      labels = c("*BCR::ABL1*", "Wild-type", "Other")
    )) %>% mutate(`*BCR::ABL1* status` := group),
    aes(x = x, y = response, colour = `Driver status`),
    alpha = 0.5
  ) +
  scale_colour_manual(values = driver.palette) +
  xlab(get_x_title(d)) +
  ylab(get_y_title(d)) +
  guides(colour = guide_legend(override.aes = list(linetype = c(rep(0, length(driver.palette)))))) + 
  new_scale_fill() +
  geom_abline(
    data = as_tibble(list(
      a = c(95.6),
      b = c(17.0),
      `Reference trend` = c("Williams et al. 17bp/year")
    )),
    aes(
      intercept = a,
      slope = b,
      fill  = `Reference trend`
    ),
    lty = 2,
    show.legend = T
  )  +
  ggtitle(unlist(str_split(
    as.character(SNV.lmer2_3.REML@call), pattern = "\\="
  ))[2]) +
  theme_bw(base_size = 16) +
  theme(legend.title = element_markdown(), legend.text = element_markdown()) 

Figure_3a_v2  
ggsave(filename = paste0("../figures/v1/", "figure_3a_V1.pdf"), plot = Figure_3a_v2, width = 9, height = 5.625 )

```

### Figure 3f

Per sample dot plot for PD51635, describing mean C\>T at CpG proportion (excluding *BCR::ABL1* clade trunk in *BCR::ABL1*-positive colonies) by sample *BCR::ABL1* fusion status, faceted by time of sampling and annotated by per sample driver status.

```{r}
PD51635.ctcpg_frq <-
  per_sample_statistics.n799.df %>%
  filter(Patient == "PD51635") %>%
  mutate(ctcpg_frq = case_when(BCR_ABL1 == "Mt" ~ ((nsub_ctcpg_adj - n_C_T_Trunk) / (nsub_adj - n_Trunk_total)
  ),
  TRUE ~ (nsub_ctcpg_adj / nsub_adj))) 

PD51635.ctcpg_frq %>% mutate(age_at_sample_exact = as.character(age_at_sample_exact)) %>% count(age_at_sample_exact, BCR_ABL1) %>% addTotal() %>%  tab_df(col.header = c("Age at sample timepoint", "BCR::ABL1 fusion status", "n"))
```

```{r, fig.height = 4.5, fig.width = 4.5, out.width="50%"}

Figure_3f.dat <-
  PD51635.ctcpg_frq %>%
  mutate(
    group = paste0(Patient, BCR_ABL1),
    # Age = as.character(age_at_sample_exact),
    Patient = factor(
      Patient,
      levels = c(
        "PD57334",
        "PD56961",
        "PD57335",
        "PD51634",
        "PD51633",
        "PD51632",
        "PD51635",
        "PD57333",
        "PD57332"
      )
    ),
    `*BCR::ABL1* status` = factor(
      BCR_ABL1,
      levels = c("Mt", "Wt"),
      labels = c("+", "-")
    ),
     `Driver status` = factor(
      driver2,
      levels = c("BCR_ABL", "Other", "WT"),
      labels = c("*BCR::ABL1*", "Other", "Wild-type")
    ),
    `C>T at CpG proportion` = ctcpg_frq,
    `Mutations SNV (adjusted)` = nsub_adj,
    `Age (years)` = age_at_sample_exact
  ) 
  
Figure_3f <-
Figure_3f.dat %>%   
  ggplot(aes( `*BCR::ABL1* status`,
             `C>T at CpG proportion`,
             colour = `Driver status`,
             group.by = group,)) +
  geom_quasirandom(width = 0.4, alpha = 0.5) +
  scale_colour_manual(values = driver.palette) +
  theme_bw(base_size = 16) +
  theme(
    legend.title = element_markdown(),
    legend.text = element_markdown(),
    axis.title.x = element_markdown(),
  ) +
  facet_nested(~ Patient + `Age (years)`, scales = "free_x") +
  force_panelsizes(cols = c(1, 2)) + theme(panel.spacing = unit(0, "lines"))

Figure_3f
ggsave(filename = paste0("../figures/v1/", "figure_3f_V1.pdf"), plot = Figure_3f, width = 4.5, height =4.5 )

```

## SessionInfo
```{r}
sessionInfo()
```