---
title: "A Simulated HSC population of an individual with CML"
output: html_document
date: "2024-10-01"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r include=FALSE}
library(rsimpop)
library(cloneRate)
library(ggplot2)

source("PHYLOFIT.R")
run_benchmark_sim=function(S=0.4,N=5e4,divrate_per_year=1,nyears_driver_acquisition = 10,nyears = 33,minprop=0.05){
  run_selection_sim(0.1,## A rapid initial division rate (death rate=0) for embryogenesis. Precise value is not important in this application.
                    final_division_rate = divrate_per_year/365,## symmetric division rate per day.
                    target_pop_size = N, ## Carrying capacity. Once this is hit the aggregate death rate matches the aggregate
                    nyears_driver_acquisition = nyears_driver_acquisition,
                    nyears = nyears,
                    fitness=log(1+S)/(divrate_per_year), ## Mutant division rate is (1+fitness)*division rate. 
                    minprop = minprop)
}
get_subsampled_tree_fixed_prop=function(selsim,N=100,nmut=10){
    mutants=which(selsim$edge[,2]<=length(selsim$tip.label) & selsim$driverid>0 & selsim$state>0)
    wt=which(selsim$edge[,2]<=length(selsim$tip.label) & selsim$driverid==0 & selsim$state>0)
    outgroup=which(selsim$edge[,2]<=length(selsim$tip.label) & selsim$state==0)
    tips=selsim$edge[c(sample(mutants,nmut,replace = FALSE),sample(wt,N-nmut,replace=FALSE),outgroup),2]
    get_subsampled_tree(selsim,-1,tips)
}
get_subtree_simple=function(selsim,N,nmut){
  st=get_subsampled_tree_fixed_prop(selsim,N=N,nmut = nmut)
  st=get_elapsed_time_tree(st)
  st$edge.length=st$edge.length/365 ## Show in years.
  list(st=st)
}
do_subsample_inference_and_plot_tree=function(selsim,nmut,maxtm=max(selsim$timestamp)/365){
  st=get_subtree_simple(selsim,nmut,nmut)
  justmutcladetree=drop.tip(st$st,st$st$tip.label[st$st$edge[st$st$state==0,2]]) ## drops the outgroup.
  ## run phylofit
  phylof=fit_clade(st$st,node = st$st$events %>% filter(value==1 & driverid==1) %>% pull(node),maxt = maxtm,nmutcolony = -1,nwtcolony = -1,maxSYear = exp(30)-1,refresh=FALSE)
  ##
  res=summary(phylof$res)$summary["s",c("2.5%","50%","97.5%")]
  bd=birthDeathMCMC(justmutcladetree,maxGrowthRate = 30,nChains = 4,nCores = 4,chainLength = 2500)
  txt1=sprintf("Phylofit       :s_est=%3.1f(%3.1f - %3.1f)",res[2],res[1],res[3])
  txt2=with(maxLikelihood(justmutcladetree),sprintf("maxLikelihood :s_est=%3.1f(%3.1f - %3.1f)",estimate,lowerBound,upperBound))
  txt3=with(bd,sprintf("birthDeathMCMC:s_est=%3.1f(%3.1f - %3.1f)",estimate,lowerBound,upperBound))
  plot_tree(justmutcladetree,cex.label=0,mar = c(1,1,4,3)+0.1)
  info=sprintf("%s\n%s\n%s",txt1,txt2,txt3)
  title(info)
  info
}
```

## Simulated history of a hypothetical CML case.

Here we simulate an individual patient with a rapidly growing clone, HSC population saturation and then TKI. 

### Simulation of rapidly growing clone (s=11)

```{r}
sid=12345379
zz=rsimpop::initSimPop(sid,bForce = TRUE)## Set seed.
s=11
S=exp(s)-1
N=1e5
nmut=60
cat("S=",S,"\nS=",sprintf("%5.1f%%",S*100),"\ns=",s,"\n")
dpy=1
#set nyear such that 99.99% of the time the pure birthdeath would have a clone size=N.  
##. N ~ exp(st)*Exp(rate=s/(1+s))
## Solve cumulative distribution=p: 1-exp(rate*N*exp(-st))=1-p
gett=function(s,N,p=0.99) -(1/s)*log(-log(p)/(N*(s/(1+s))))
nyd=5## Year that the driver is acquired.  We set 5 years to avoid unnecessarily long simulation.
nyear1=nyd+gett(s,N,p=0.9)
nyear2=nyd+gett(s,N,p=0.9999)+ 3 ## Add 3 years to make the time at saturation longer.
```
```{r results='hide'}
selsim=run_benchmark_sim(S=S,divrate_per_year = dpy,N = N,nyears_driver_acquisition = nyd,nyear=nyear1,minprop = 0.5)
ggplot(data=data.frame(ts=selsim$timestamp/365,mutpop=selsim$totaldrivercount),aes(x=ts,y=mutpop))+geom_line()+
  xlim(c(nyd-1,max(selsim$timestamp/365)+1))+xlab("Time (Years)")+ylab("Mutant Population Size")+ theme_bw()
```

### Sampled trees and inferred growth rates

We only have one individual and take 9 hypothetical samples, each resulting in 60 colonies.  This might be the case for an incidental diagnosis.

The inferred growth rates are generally quite accurate, reflecting the fact that expansion has been largely pure exponential growth

```{r, fig.width=14,fig.height=12,results='hide'}
par(mfcol=c(3,3))
for(i in 1:9){
  zz=do_subsample_inference_and_plot_tree(selsim,nmut)
}
```

### Evolve clone at saturation population for about 3 years.

Note that longer the periods of saturation will result in greater inaccuracy in methods that assume pure exponential growth.   

```{r}
params = selsim$params
params[["n_sim_days"]] = nyear2*365
params[["maxt"]] = NULL
selsim2 = sim_pop(selsim, params = params, selsim$cfg)
ggplot(data=data.frame(ts=selsim2$timestamp/365,mutpop=selsim2$totaldrivercount),aes(x=ts,y=mutpop))+geom_line()+
  xlim(c(nyd-1,max(selsim2$timestamp/365)+1))+xlab("Time (Years)")+ylab("Mutant Population Size")+ theme_bw()
```

Again take hypothetical samples.   The estimated rates are very heterogeneous, and sometimes too low depending on whether late coalescences are captured:

```{r, fig.width=14,fig.height=12,results='hide'}
par(mfcol=c(3,3))
for(i in 1:9){
  zz=do_subsample_inference_and_plot_tree(selsim2,nmut)
}
```

## Now apply TKI 

Note that in the following we assume that TKI works by apoptosis which manifests as an increase in death rate, but division rate is maintained.   

Assuming that MMR (x% clone size) is attained in time t (in years) then the corresponding growth rate "s" (actually negative representing contraction) is found by: $e^{s t}=x$ giving $s=\log(x)/t$. In the following we follow the example of PD57335 where at colony sampling the mutant fraction is around 20% after 0.6 years of TKI.   
```{r}
post.treatment.years=0.6
oselsim=selsim
selsim=selsim2
## Put the driver in a separate compartment and specify the rate.  
cfg=addCellCompartment(selsim$cfg,population=1e6,
                       rate=(1+selsim$cfg$info %>% filter(id==1 & val==1) %>% pull("fitness"))*
                         selsim$cfg$compartment %>% filter(val==1) %>% pull("rate"),
                       descr="Post-TKI-Driver")
idx=which(selsim$driverid==1 & selsim$edge[,2]<=Ntip(selsim)) ## select drivers

selsim$state[idx]=2 #' switch to the new comparement (val=2)
selsim$events=rbind(
  selsim$events,
  data.frame(value=2,driverid=1,node=selsim$edge[idx,2],ts=max(selsim$timestamp),uid=max(selsim$events$uid)+1:length(idx))
) ## Update the events
selsim$currentEventUID=max(selsim$event$uid)
cfg$compartment$death_rate[3]=cfg$compartment$rate[3]-((1/365)*(log(0.2)/0.6))## MMR within 12 months
cfg
idx1=which(cfg$info$val==1 & cfg$info$id==1)
idx2=which(cfg$info$val==2)
cfg$info$population[idx2]=cfg$info$population[idx1]
cfg$info$population[idx1]=0
cfg$info$id[idx2]=1
print(cfg)
params = selsim$params
params[["n_sim_days"]] = ceiling(max(selsim$timestamp)+post.treatment.years * 365)
params[["maxt"]] = NULL
```
```{r results='hide'}
selsim_tki = sim_pop(selsim %>% (function(x) {x$cfg=cfg;x}), params = params, cfg)
selsim=oselsim ## 
```

The mutant population has shrunk:

```{r}
g1=ggplot(data=data.frame(ts=selsim_tki$timestamp/365,mutpop=selsim_tki$totaldrivercount),aes(x=ts,y=mutpop))+geom_line()+scale_x_continuous(limits=c(0,ceiling(max(selsim_tki$timestamp/365))),breaks=0:ceiling(max(selsim_tki$timestamp/365)))+xlab("Time (Years)")+ylab("Mutant Population Size")+ theme_bw()
g1=g1+annotate("rect",xmin=nyd,xmax=nyear1,ymin=-Inf,ymax=Inf,fill="grey",alpha=0.2)
g1=g1+annotate("rect",xmin=nyear1,xmax=nyear2,ymin=-Inf,ymax=Inf,fill="blue",alpha=0.2)
g1=g1+annotate("rect",xmin=nyear2,xmax=max(selsim_tki$timestamp/365),ymin=-Inf,ymax=Inf,fill="red",alpha=0.2)
print(g1)
```

Note that the sampled trees are exhibiting more late coalescences, resulting in much lower estimates of s for both birthDeathMCMC and maxLikelihood, but not for phylofit where the estimates are relatively stable.

```{r, fig.width=14,fig.height=12,results='hide'}
saveRDS(selsim_tki,"selsimTKI.RDS")
par(mfcol=c(3,3))
for(i in 1:9){
  zz=do_subsample_inference_and_plot_tree(selsim_tki,nmut)
}
```

## Conclusions

It seems that Phylofit inferences of the early growth rate are relatively robust to later perturbations to mutant population growth trajectory.   





